---
title: "FA 2"
author: "Awit, Julia Theressa and Deloyola, Jhen Reyniel L."
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


```{r}
library(tidyverse)
```

Perform a data analysis to study trends in tuberculosis (TB) cases worldwide over time. The two relevant datasets are:

- **who**: Information about TB cases in various countries from 1980 to 2013 (Source: 2014 WHO Global Tuberculosis Report) - built into the tidyr package.
- **population.csv**: Population data of each country across time (Source: The World Bank).

# 2. Import


## 2.1 Preview the contents of who and population.csv by inspecting the files
```{r}


# Load the built-in WHO TB dataset from tidyr
# Population file (main World Bank CSV)
pop_file <- "main.csv"

# Check if file exists
if (!file.exists(pop_file)) {
  stop("main.csv not found. Make sure it is in the same directory as the Rmd file.")
}

# Preview first 10 lines
cat("Preview of population main.csv:\n")
print(readLines(pop_file, n = 10))

```

## 2.2 Import the data into tibbles named who and population
```{r}
# Import population data (skip metadata rows)
population <- read_csv(pop_file, skip = 4, show_col_types = FALSE)

cat("Population dataset imported successfully!\n")

```

## 2.3 Determine the number of rows and columns in each tibble
```{r}
cat("Population dataset:\n")
cat("Rows:", nrow(population), "\n")
cat("Columns:", ncol(population), "\n")

```

## 2.4 Check the summary of variable types for population.csv. Fix any anomalies and store the corrected data in population2
```{r}
cat("Summary of population dataset BEFORE fixing:\n")
str(population)

# Convert year columns to numeric
population2 <- population %>%
  mutate(across(matches("^\\d{4}$"), as.numeric))

cat("\nSummary of corrected population dataset (population2):\n")
str(population2)

```

# 3. Tidy Data

## 3.1 who Dataset

### Description of Columns:

- **country**: Country name.
- **iso2**: Two-digit country code.
- **iso3**: Three-digit country code.
- **year**: Year.
- Variables like **new_ep_f014**:
  - **ep**: TB type (e.g., rel = relapse, ep = extrapulmonary).
  - **f**: Sex (e.g., f = female).
  - **014**: Age group (e.g., 0-14 years).

### Steps:

#### 3.1.1 Identify the variables in the dataset
```{r}
cat("Variables in WHO dataset:\n")
names(who)
```

#### 3.1.2 Perform a pivot operation to make the data tidy, storing the result in who2
```{r}
# Pivot longer: convert all the "new_" columns into key-value pairs
who2 <- who %>%
  pivot_longer(
    cols = matches("new_"),
    names_to = "key",
    values_to = "cases"
  )

cat("WHO2 dataset (after pivot):\n")
head(who2, 10)
```

#### 3.1.3 Separate values like new_ep_f014 into components (e.g., new, ep, f014). Remove the column containing new, and store the result in who3
```{r}
# Separate the key column and remove "new"
who3 <- who2 %>%
  separate(key, into = c("new", "type", "sexage"), sep = "_") %>%
  select(-new)

cat("WHO3 dataset (after separating and removing 'new'):\n")
head(who3, 10)
```

#### 3.1.4 Further separate values like f014 into f and 014, storing the result in who_tidy
```{r}
# Separate sexage into sex and age
who_tidy <- who3 %>%
  separate(sexage, into = c("sex", "age"), sep = 1)

cat("WHO_tidy dataset (final tidy format):\n")
head(who_tidy, 10)
cat("\nDimensions:", nrow(who_tidy), "rows x", ncol(who_tidy), "columns\n")
```

## 3.2 Population Dataset

#### 3.2.1 Identify the variables in this dataset
```{r}
cat("Variables in Population dataset:\n")
names(population2)
```

#### 3.2.2 Perform a pivot operation to tidy the data, storing the result in population3
```{r}
# Select relevant columns and pivot year columns into long format
population3 <- population2 %>%
  select(`Country Name`, `Country Code`, matches("^\\d{4}$")) %>%
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "population"
  )

cat("Population3 dataset (after pivot):\n")
head(population3, 10)
```

#### 3.2.3 Cast the population variable to an appropriate data type, storing the result in population_tidy
```{r}
# Convert year to integer and ensure population is numeric
population_tidy <- population3 %>%
  mutate(
    year = as.integer(year),
    population = as.numeric(population)
  )

cat("Population_tidy dataset (with correct data types):\n")
head(population_tidy, 10)

cat("\nData types:\n")
str(population_tidy)
```

## 3.3 Join Datasets

#### 3.3.1 Identify the variable(s) required to join who_tidy and population_tidy
```{r}
cat("Key variables for joining:\n")
cat("WHO dataset columns:", paste(names(who_tidy), collapse = ", "), "\n")
cat("Population dataset columns:", paste(names(population_tidy), collapse = ", "), "\n\n")

cat("Common keys:\n")
cat("- Country code: 'iso3' in WHO, 'Country Code' in Population\n")
cat("- Year: 'year' in both datasets\n")
```

#### 3.3.2 Rename columns as needed to align variable names between datasets
```{r}
# Rename Population dataset columns to match WHO dataset
population_tidy <- population_tidy %>%
  rename(
    country = `Country Name`,
    iso3 = `Country Code`
  )

cat("Population_tidy dataset (after renaming):\n")
head(population_tidy, 10)
```

#### 3.3.3 Join the datasets into a tibble called tuberculosis
```{r}
# Left join: keep all WHO records and add population data where available
tuberculosis <- who_tidy %>%
  left_join(population_tidy, by = c("iso3", "year"))

cat("Tuberculosis dataset (after join):\n")
head(tuberculosis, 10)
cat("\nDimensions:", nrow(tuberculosis), "rows x", ncol(tuberculosis), "columns\n")
```

## 3.4 Clean Up Data

#### 3.4.1 Remove unnecessary variables from tuberculosis
```{r}
# Keep only essential variables
tuberculosis <- tuberculosis %>%
  select(country.x, iso2, iso3, year, type, sex, age, cases, population) %>%
  rename(country = country.x)

cat("Cleaned tuberculosis dataset (essential variables only):\n")
head(tuberculosis, 10)
```

#### 3.4.2 Filter out rows with NA values
```{r}
# Remove rows where cases or population is NA
cat("Before filtering NA values:\n")
cat("Rows:", nrow(tuberculosis), "\n")

tuberculosis <- tuberculosis %>%
  filter(!is.na(cases), !is.na(population))

cat("\nAfter filtering NA values:\n")
cat("Rows:", nrow(tuberculosis), "\n")
cat("Columns:", ncol(tuberculosis), "\n")
```

#### 3.4.3 Save the cleaned data back into tuberculosis
```{r}
# Data is already saved in the tuberculosis tibble
cat("Final cleaned tuberculosis dataset:\n")
head(tuberculosis, 10)
cat("\nFinal dimensions:", nrow(tuberculosis), "rows x", ncol(tuberculosis), "columns\n")
```

# 4. Data Manipulation

## 4.1 Determine the total TB cases among men and women in the 21st century in the United States. Identify which sex had more cases.
```{r}
# Filter for United States, 21st century (2000-2013 in our dataset)
us_tb_21st <- tuberculosis %>%
  filter(country == "United States of America", year >= 2000, year <= 2013) %>%
  group_by(sex) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE)) %>%
  arrange(desc(total_cases))

cat("Total TB cases in the United States (21st century, 2000-2013):\n")
print(us_tb_21st)

cat("\n*** RESULT: Sex with more cases: **", us_tb_21st$sex[1], "** with **", 
    format(us_tb_21st$total_cases[1], big.mark = ","), "** cases ***\n")
```

## 4.2 Create a new variable, cases_per_100k, representing TB cases per 100,000 people by year, sex, age group, and TB type
```{r}
# Calculate cases per 100,000 population
tuberculosis <- tuberculosis %>%
  mutate(cases_per_100k = (cases / population) * 100000)

cat("Tuberculosis dataset with cases_per_100k:\n")
head(tuberculosis, 10)

# Show summary statistics
cat("\nSummary of cases_per_100k:\n")
summary(tuberculosis$cases_per_100k)
```

## 4.3 Identify:

### The country and year with the highest cases per 100k
```{r}
highest_cases <- tuberculosis %>%
  filter(!is.na(cases_per_100k)) %>%
  arrange(desc(cases_per_100k)) %>%
  slice(1) %>%
  select(country, year, type, sex, age, cases, population, cases_per_100k)

cat("*** HIGHEST CASES PER 100k: ***\n")
print(highest_cases)
```

### The country and year with the lowest cases per 100k
```{r}
lowest_cases <- tuberculosis %>%
  filter(!is.na(cases_per_100k), cases_per_100k > 0) %>%
  arrange(cases_per_100k) %>%
  slice(1) %>%
  select(country, year, type, sex, age, cases, population, cases_per_100k)

cat("*** LOWEST CASES PER 100k (excluding zero): ***\n")
print(lowest_cases)
```

# 5. Data Visualization

## 5.1 Plot the total cases per 100k as a function of year for China, India, and the United States
```{r fig.width=10, fig.height=6}
# Aggregate data for the three countries
plot_data_5_1 <- tuberculosis %>%
  filter(country %in% c("China", "India", "United States of America")) %>%
  group_by(country, year) %>%
  summarise(total_cases_per_100k = sum(cases_per_100k, na.rm = TRUE), .groups = "drop")

# Create the plot with log scale on y-axis
ggplot(plot_data_5_1, aes(x = year, y = total_cases_per_100k, color = country)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Total TB Cases per 100k Over Time",
    subtitle = "China, India, and United States (1980-2013) - Log Scale",
    x = "Year",
    y = "Total Cases per 100k (log scale)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  ) +
  scale_color_manual(values = c("China" = "#E41A1C", 
                                 "India" = "#4DAF4A", 
                                 "United States of America" = "#377EB8"))
```

### Emerging Patterns:

Based on the visualization above, several patterns emerge:

1. **India** consistently shows the highest TB case rates per 100,000 population throughout the entire period (1980-2013), though there is a gradual declining trend, particularly after 2000.

2. **China** demonstrates a significant and steady decreasing trend in TB cases per 100k over the time period. The decline is particularly pronounced from 1980 through the early 2000s, showing successful TB control measures.

3. **United States** maintains the lowest TB case rates among the three countries, with rates significantly below both China and India. The US shows a relatively stable, low rate with a gentle decline over time.

4. **All three countries** show downward trends overall, indicating global improvements in TB prevention, detection, and treatment programs during this period.

5. The **gap between countries** narrows somewhat over time, particularly as China's rates decline more rapidly, though India maintains notably higher rates throughout.

## 5.2 Compare distributions of total cases per 100k (summed over years, sexes, and TB types) across age groups
```{r fig.width=10, fig.height=6}
# Aggregate data by age group
plot_data_5_2 <- tuberculosis %>%
  group_by(age) %>%
  summarise(total_cases_per_100k = sum(cases_per_100k, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_cases_per_100k))

# Create the bar plot with log scale
ggplot(plot_data_5_2, aes(x = reorder(age, total_cases_per_100k), y = total_cases_per_100k)) +
  geom_col(fill = "pink", alpha = 0.8) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Distribution of Total TB Cases per 100k Across Age Groups",
    subtitle = "Summed over years, sexes, and TB types (1980-2013) - Log Scale",
    x = "Age Group",
    y = "Total Cases per 100k (log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  ) +
  coord_flip()
```

### Observed Patterns:

The bar chart reveals important epidemiological patterns in TB distribution across age groups:

1. **Working-age adults** (particularly the 35-44, 45-54, and 25-34 age groups) show the highest total TB case rates per 100k. This likely reflects greater exposure risk due to social and occupational factors.

2. **Older adults** (55-64 and 65+ age groups) also show elevated TB case rates, which may be due to weaker immune systems and reactivation of latent TB infections acquired earlier in life.

3. **Young adults** (15-24 years) have moderate TB rates, reflecting both new infections and the beginning of reactivation cases.

4. **Children and adolescents** (0-14 age group) show relatively lower case rates compared to adults, which is typical for TB epidemiology as children are less likely to develop active disease, though they remain vulnerable.

5. The pattern follows a **characteristic age distribution** for TB, with peak incidence in economically productive age groups (25-64 years), reflecting both higher transmission risk and the impact of TB on workforce populations.

## 5.3 Create a plot to evaluate whether the number of cases per 100k in 2000 was related to a country's population
```{r fig.width=10, fig.height=6}
# Filter data for year 2000 and aggregate by country
plot_data_5_3 <- tuberculosis %>%
  filter(year == 2000) %>%
  group_by(country, population) %>%
  summarise(total_cases_per_100k = sum(cases_per_100k, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(population), !is.na(total_cases_per_100k), population > 0)

# Create the scatter plot with both axes on log scale
ggplot(plot_data_5_3, aes(x = population, y = total_cases_per_100k)) +
  geom_point(alpha = 0.6, color = "red", size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue", linewidth = 1.2) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Relationship Between TB Cases per 100k and Population in 2000",
    subtitle = "Each point represents a country (both axes on log scale)",
    x = "Population (log scale)",
    y = "Cases per 100k (log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

# Calculate correlation
correlation <- cor(log10(plot_data_5_3$population), 
                   log10(plot_data_5_3$total_cases_per_100k), 
                   use = "complete.obs")
cat("\nCorrelation between log(population) and log(cases per 100k):", round(correlation, 3), "\n")
```

### Conclusion:

The analysis shows that a country’s population size does not strongly relate to its TB case rate per 100,000 people in 2000. The scatter plot and near-zero correlation indicate that similar populations can have very different TB rates, and vice versa. This makes sense because TB rates are already adjusted for population size, so the total population alone isn’t a good predictor. Other factors like healthcare access, poverty, public health programs, and HIV prevalence have a much bigger impact on TB rates. Therefore, TB control efforts should focus on local conditions and risk factors rather than population size.