---
title: "SA1 Exploratory Analysis"
author: "Awit, Julia Theressa and Deloyola, Jhen reyniel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
```

---

## 1.1 Load Dataset & Summarize Structure

```{r load-libraries}
library(tidyverse)   # Core data wrangling + ggplot2
library(moments)     # Skewness and kurtosis
library(gridExtra)   # Multi-panel plots
library(scales)      # Axis formatting
library(MASS)        # Robust regression (rlm)
library(e1071)       # Mode helper
library(ggcorrplot)  # Correlation plot
library(car)         # VIF for regression

set.seed(42)
```

```{r load-data}
df <- read.csv("EDA_Ecommerce_Assessment.csv", stringsAsFactors = FALSE)

# Clean factor columns
df <- df %>%
  mutate(
    Gender            = factor(trimws(Gender)),
    Category          = factor(trimws(Category)),
    Satisfaction_Score = as.integer(Satisfaction_Score)
  )

cat("=== Dataset Dimensions ===\n")
cat("Rows:", nrow(df), " | Columns:", ncol(df), "\n\n")

cat("=== Column Types ===\n")
str(df)
```

```{r data-overview}
cat("=== Summary Statistics ===\n")
summary(df)

cat("\n=== Missing Values ===\n")
mv <- colSums(is.na(df))
print(mv)
if (sum(mv) == 0) cat("No missing values found.\n")

cat("\n=== Category Distribution ===\n")
print(table(df$Category))

cat("\n=== Gender Distribution ===\n")
print(table(df$Gender))
```

> **Dataset Overview:**  
> The dataset contains **3,000 customer transactions** across 10 variables covering demographics
> (Gender, Age), browsing behaviour (Browsing_Time), purchasing metrics (Purchase_Amount,
> Number_of_Items, Discount_Applied, Total_Transactions), product Category, and
> Satisfaction_Score. There are no missing values. Categories are evenly distributed
> (~575–630 records each): Beauty, Books, Clothing, Electronics, and Home & Kitchen.

---

## 1.2 Histograms & Boxplots

### Purchase_Amount

```{r hist-box-purchase}
p1 <- ggplot(df, aes(x = Purchase_Amount)) +
  geom_histogram(bins = 35, fill = "#3498DB", color = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = mean(Purchase_Amount)), color = "#E74C3C",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Distribution of Purchase Amount",
       subtitle = "Red dashed line = mean",
       x = "Purchase Amount (USD)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(df, aes(y = Purchase_Amount)) +
  geom_boxplot(fill = "#3498DB", alpha = 0.7, outlier.colour = "#E74C3C",
               outlier.size = 1.5) +
  labs(title = "Boxplot of Purchase Amount", y = "Purchase Amount (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())

grid.arrange(p1, p2, ncol = 2)
```

### Number_of_Items

```{r hist-box-items}
p3 <- ggplot(df, aes(x = Number_of_Items)) +
  geom_histogram(bins = 10, fill = "#2ECC71", color = "white", alpha = 0.85) +
  labs(title = "Distribution of Number of Items",
       x = "Number of Items", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p4 <- ggplot(df, aes(y = Number_of_Items)) +
  geom_boxplot(fill = "#2ECC71", alpha = 0.7, outlier.colour = "#E74C3C",
               outlier.size = 1.5) +
  labs(title = "Boxplot of Number of Items", y = "Number of Items") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())

grid.arrange(p3, p4, ncol = 2)
```

### Satisfaction_Score

```{r hist-box-satisfaction}
p5 <- ggplot(df, aes(x = factor(Satisfaction_Score))) +
  geom_bar(fill = "#9B59B6", color = "white", alpha = 0.85) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.4, size = 4) +
  labs(title = "Distribution of Satisfaction Score",
       x = "Satisfaction Score (1–5)", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p6 <- ggplot(df, aes(y = Satisfaction_Score)) +
  geom_boxplot(fill = "#9B59B6", alpha = 0.7) +
  labs(title = "Boxplot of Satisfaction Score", y = "Satisfaction Score") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())

grid.arrange(p5, p6, ncol = 2)
```

> **Insights:**  
> - **Purchase_Amount** is broadly spread across ~$0–$500 with a roughly uniform/slight right skew.  
> - **Number_of_Items** ranges from 1–10 and appears approximately uniform (consistent basket sizes).  
> - **Satisfaction_Score** is nearly uniformly distributed across 1–5, indicating no overwhelming satisfaction bias in this dataset.

---

## 1.3 Measures of Central Tendency & Spread for Purchase_Amount

```{r central-tendency}
# Mode helper function
get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

pa <- df$Purchase_Amount

cat("=== Central Tendency ===\n")
cat("Mean:   ", round(mean(pa), 2), "\n")
cat("Median: ", round(median(pa), 2), "\n")
cat("Mode:   ", round(get_mode(pa), 2), "\n")

cat("\n=== Spread ===\n")
cat("Variance:           ", round(var(pa), 2), "\n")
cat("Std Deviation:      ", round(sd(pa), 2), "\n")
cat("IQR:                ", round(IQR(pa), 2), "\n")
cat("Range:              ", round(diff(range(pa)), 2), "\n")
cat("Skewness:           ", round(skewness(pa), 4), "\n")
cat("Kurtosis:           ", round(kurtosis(pa), 4), "\n")
```

> **Interpretation:**  
> The mean and median of Purchase_Amount are close, indicating a near-symmetric distribution.
> The standard deviation is large relative to the mean, reflecting wide variability in spending.
> Skewness close to 0 confirms approximate symmetry; kurtosis near 3 suggests a near-normal tail shape.

---

## 1.4 Density Plots by Gender

```{r density-gender}
# Browsing Time by Gender
p_bt <- ggplot(df, aes(x = Browsing_Time, fill = Gender)) +
  geom_density(alpha = 0.50, linewidth = 0.8) +
  scale_fill_manual(values = c("Female" = "#E74C3C", "Male" = "#3498DB")) +
  labs(title = "Browsing Time Distribution by Gender",
       x = "Browsing Time (min)", y = "Density", fill = "Gender") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

# Purchase Amount by Gender
p_pa <- ggplot(df, aes(x = Purchase_Amount, fill = Gender)) +
  geom_density(alpha = 0.50, linewidth = 0.8) +
  scale_fill_manual(values = c("Female" = "#E74C3C", "Male" = "#3498DB")) +
  labs(title = "Purchase Amount Distribution by Gender",
       x = "Purchase Amount (USD)", y = "Density", fill = "Gender") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

grid.arrange(p_bt, p_pa, ncol = 2)
```

```{r gender-summary}
df %>%
  group_by(Gender) %>%
  summarise(
    Mean_Browsing  = round(mean(Browsing_Time), 2),
    Median_Browsing = round(median(Browsing_Time), 2),
    Mean_Purchase  = round(mean(Purchase_Amount), 2),
    Median_Purchase = round(median(Purchase_Amount), 2),
    .groups = "drop"
  )
```

> **Insight:**  
> Both gender groups show very similar distributions for Browsing_Time and Purchase_Amount,
> suggesting gender does not strongly differentiate browsing or spending behavior in this dataset.

---

## 1.5 Logarithmic & Square Root Transformation on Browsing_Time

```{r transformations}
df <- df %>%
  mutate(
    BT_log  = log1p(Browsing_Time),   # log(1+x) handles any near-zero values
    BT_sqrt = sqrt(Browsing_Time)
  )

cat("=== Skewness Comparison ===\n")
cat("Original  Browsing_Time skewness:", round(skewness(df$Browsing_Time), 4), "\n")
cat("Log-transformed       skewness:", round(skewness(df$BT_log), 4), "\n")
cat("Sqrt-transformed      skewness:", round(skewness(df$BT_sqrt), 4), "\n")
```

```{r transform-plots}
t1 <- ggplot(df, aes(x = Browsing_Time)) +
  geom_histogram(bins = 30, fill = "#E67E22", color = "white", alpha = 0.85) +
  labs(title = "Original Browsing Time",
       subtitle = paste("Skewness:", round(skewness(df$Browsing_Time), 3)),
       x = "Browsing Time (min)", y = "Count") +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold"))

t2 <- ggplot(df, aes(x = BT_log)) +
  geom_histogram(bins = 30, fill = "#27AE60", color = "white", alpha = 0.85) +
  labs(title = "Log-Transformed Browsing Time",
       subtitle = paste("Skewness:", round(skewness(df$BT_log), 3)),
       x = "log(Browsing Time)", y = "Count") +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold"))

t3 <- ggplot(df, aes(x = BT_sqrt)) +
  geom_histogram(bins = 30, fill = "#8E44AD", color = "white", alpha = 0.85) +
  labs(title = "Sqrt-Transformed Browsing Time",
       subtitle = paste("Skewness:", round(skewness(df$BT_sqrt), 3)),
       x = "sqrt(Browsing Time)", y = "Count") +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold"))

grid.arrange(t1, t2, t3, ncol = 3)
```

> **Evaluation:**  
> If Browsing_Time is right-skewed, both log and square root transformations reduce skewness
> toward zero, making the distribution more symmetric. The log transformation typically
> achieves a stronger correction for heavy right skew. Near-zero skewness after transformation
> indicates a more normal-like distribution, which improves the validity of regression assumptions.

---

## 1.6 Simple Linear Regression: Purchase_Amount ~ Browsing_Time

```{r slr-model}
slr <- lm(Purchase_Amount ~ Browsing_Time, data = df)

cat("=== Simple Linear Regression Summary ===\n")
summary(slr)
```

```{r slr-plot}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.25, color = "#3498DB", size = 1.2) +
  geom_smooth(method = "lm", color = "#E74C3C", linewidth = 1.2, se = TRUE,
              fill = "#F1948A") +
  labs(
    title    = "Purchase Amount vs. Browsing Time",
    subtitle = paste0("Simple Linear Regression | R² = ",
                      round(summary(slr)$r.squared, 4)),
    x = "Browsing Time (min)", y = "Purchase Amount (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

```{r slr-diagnostics}
cat("=== Model Coefficients ===\n")
cat("Intercept (β₀):", round(coef(slr)[1], 4), "\n")
cat("Slope     (β₁):", round(coef(slr)[2], 4), "\n")
cat("R-squared:     ", round(summary(slr)$r.squared, 4), "\n")
cat("Adj R-squared: ", round(summary(slr)$adj.r.squared, 4), "\n")
cat("p-value (slope):", round(summary(slr)$coefficients[2,4], 4), "\n")
```

> **Interpretation:**  
> - **Intercept (β₀):** The predicted Purchase_Amount when Browsing_Time = 0.  
> - **Slope (β₁):** For each additional minute of browsing, Purchase_Amount changes by β₁ USD.
>   A positive slope indicates more browsing associates with higher spending; negative would indicate the opposite.  
> - **R²:** The proportion of variance in Purchase_Amount explained by Browsing_Time alone.
>   A low R² suggests that browsing time alone is insufficient — other variables are needed.  
> - **p-value:** If < 0.05, Browsing_Time is a statistically significant predictor.

---

# Unit 2: Bivariate Data Analysis

## 2.1 Scatter Plot: Purchase_Amount vs. Number_of_Items

```{r scatter-items}
ggplot(df, aes(x = Number_of_Items, y = Purchase_Amount)) +
  geom_jitter(alpha = 0.25, color = "#2980B9", size = 1.5, width = 0.2) +
  geom_smooth(method = "lm", color = "#E74C3C", linewidth = 1.2,
              se = TRUE, fill = "#F1948A") +
  geom_smooth(method = "loess", color = "#27AE60", linewidth = 1,
              linetype = "dashed", se = FALSE) +
  labs(
    title    = "Purchase Amount vs. Number of Items",
    subtitle = "Red = Linear fit | Green dashed = LOESS",
    x = "Number of Items", y = "Purchase Amount (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

# Correlation
cat("Pearson correlation (Purchase_Amount ~ Number_of_Items):",
    round(cor(df$Purchase_Amount, df$Number_of_Items), 4), "\n")
```

> **Insight:**  
> The scatter plot shows the relationship between basket size and total spend.
> A positive correlation would suggest larger baskets drive higher purchase amounts.
> The LOESS curve reveals whether the relationship is truly linear or has non-linear components.

---

## 2.2 Polynomial Regression vs. Simple Linear Regression

```{r poly-regression}
# Simple linear model
lm_simple <- lm(Purchase_Amount ~ Browsing_Time, data = df)

# Polynomial degree 2
lm_poly2 <- lm(Purchase_Amount ~ poly(Browsing_Time, 2, raw = TRUE), data = df)

# Polynomial degree 3
lm_poly3 <- lm(Purchase_Amount ~ poly(Browsing_Time, 3, raw = TRUE), data = df)

cat("=== Model Comparison ===\n")
cat("Linear   R²:", round(summary(lm_simple)$r.squared, 4),
    "| AIC:", round(AIC(lm_simple), 2), "\n")
cat("Poly (2) R²:", round(summary(lm_poly2)$r.squared, 4),
    "| AIC:", round(AIC(lm_poly2), 2), "\n")
cat("Poly (3) R²:", round(summary(lm_poly3)$r.squared, 4),
    "| AIC:", round(AIC(lm_poly3), 2), "\n")

# ANOVA comparison
cat("\n=== ANOVA: Linear vs Poly(2) ===\n")
print(anova(lm_simple, lm_poly2))
cat("\n=== ANOVA: Poly(2) vs Poly(3) ===\n")
print(anova(lm_poly2, lm_poly3))
```

```{r poly-plot}
bt_seq <- seq(min(df$Browsing_Time), max(df$Browsing_Time), length.out = 300)
pred_df <- tibble(
  Browsing_Time = bt_seq,
  Linear  = predict(lm_simple, newdata = tibble(Browsing_Time = bt_seq)),
  Poly2   = predict(lm_poly2,  newdata = tibble(Browsing_Time = bt_seq)),
  Poly3   = predict(lm_poly3,  newdata = tibble(Browsing_Time = bt_seq))
) %>%
  pivot_longer(c(Linear, Poly2, Poly3), names_to = "Model", values_to = "Predicted")

ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.15, color = "gray60", size = 1) +
  geom_line(data = pred_df, aes(y = Predicted, color = Model), linewidth = 1.1) +
  scale_color_manual(values = c("Linear" = "#E74C3C",
                                "Poly2"  = "#3498DB",
                                "Poly3"  = "#2ECC71")) +
  labs(
    title    = "Polynomial Regression: Purchase Amount ~ Browsing Time",
    subtitle = "Comparing Linear, Degree-2, and Degree-3 polynomial fits",
    x = "Browsing Time (min)", y = "Purchase Amount (USD)", color = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Comparison:**  
> If ANOVA p-values are significant (< 0.05), the polynomial terms add meaningful fit over the linear model.
> A lower AIC favors a better model with less complexity penalty.
> If polynomial R² improvement is negligible, the simpler linear model is preferred (parsimony principle).

---

## 2.3 LOESS Smoothing

```{r loess-plot}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.20, color = "#95A5A6", size = 1.2) +
  geom_smooth(method = "lm",    aes(color = "Linear (OLS)"),
              linewidth = 1.1, se = FALSE) +
  geom_smooth(method = "loess", aes(color = "LOESS (span=0.75)"),
              span = 0.75, linewidth = 1.1, se = TRUE, fill = "#AED6F1") +
  geom_smooth(method = "loess", aes(color = "LOESS (span=0.3)"),
              span = 0.30, linewidth = 1.0, linetype = "dashed", se = FALSE) +
  scale_color_manual(
    values = c("Linear (OLS)"      = "#E74C3C",
               "LOESS (span=0.75)" = "#2980B9",
               "LOESS (span=0.3)"  = "#27AE60")
  ) +
  labs(
    title    = "LOESS Smoothing: Purchase Amount vs. Browsing Time",
    subtitle = "Comparing OLS with LOESS at two bandwidth levels",
    x = "Browsing Time (min)", y = "Purchase Amount (USD)", color = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Insight:**  
> LOESS with span = 0.75 provides a smooth, globally-influenced trend, revealing any broad
> non-linear pattern. Span = 0.3 is more locally adaptive — it follows local fluctuations
> more closely but can overfit. Where LOESS diverges noticeably from OLS, it signals
> true non-linearity that the linear model misses.

---

## 2.4 Robust Regression vs. OLS

```{r robust-regression}
# OLS
ols_model <- lm(Purchase_Amount ~ Browsing_Time, data = df)

# Huber robust regression (MASS::rlm)
huber_model <- rlm(Purchase_Amount ~ Browsing_Time, data = df,
                   method = "M", psi = psi.huber)

# Bisquare (Tukey) robust regression
tukey_model <- rlm(Purchase_Amount ~ Browsing_Time, data = df,
                   method = "M", psi = psi.bisquare)

cat("=== Coefficient Comparison ===\n")
coef_table <- tibble(
  Method    = c("OLS", "Huber", "Tukey (Bisquare)"),
  Intercept = c(round(coef(ols_model)[1],   2),
                round(coef(huber_model)[1], 2),
                round(coef(tukey_model)[1], 2)),
  Slope     = c(round(coef(ols_model)[2],   4),
                round(coef(huber_model)[2], 4),
                round(coef(tukey_model)[2], 4))
)
print(coef_table)
```

```{r robust-plot}
bt_grid <- tibble(Browsing_Time = seq(min(df$Browsing_Time),
                                      max(df$Browsing_Time), length.out = 300))

pred_robust <- bt_grid %>%
  mutate(
    OLS   = predict(ols_model,   newdata = bt_grid),
    Huber = predict(huber_model, newdata = bt_grid),
    Tukey = predict(tukey_model, newdata = bt_grid)
  ) %>%
  pivot_longer(c(OLS, Huber, Tukey), names_to = "Method", values_to = "Predicted")

ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.15, color = "gray70", size = 1) +
  geom_line(data = pred_robust,
            aes(y = Predicted, color = Method, linetype = Method),
            linewidth = 1.2) +
  scale_color_manual(values = c("OLS"   = "#E74C3C",
                                "Huber" = "#3498DB",
                                "Tukey" = "#27AE60")) +
  scale_linetype_manual(values = c("OLS" = "solid",
                                   "Huber" = "dashed",
                                   "Tukey" = "dotdash")) +
  labs(
    title    = "Robust Regression vs. OLS",
    subtitle = "Comparing OLS, Huber M-estimator, and Tukey Bisquare",
    x = "Browsing Time (min)", y = "Purchase Amount (USD)",
    color = "Method", linetype = "Method"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Comparison:**  
> OLS minimizes squared residuals, making it sensitive to outliers — extreme purchase values
> can heavily pull the regression line. Huber regression down-weights large residuals
> progressively, while Tukey bisquare completely ignores observations beyond a threshold.
> If the lines are similar, outliers are not severely affecting OLS. If they diverge,
> robust regression provides a more trustworthy estimate of the true relationship.

---

# Unit 3: Trivariate / Hypervariate Data Analysis

## 3.1 Interaction Effects: Browsing_Time × Category on Purchase_Amount

```{r interaction-model}
# Fit interaction model
int_model <- lm(Purchase_Amount ~ Browsing_Time * Category, data = df)

cat("=== Interaction Model Summary ===\n")
summary(int_model)
```

```{r interaction-plot}
# Predict over Browsing_Time range for each Category
bt_range <- seq(min(df$Browsing_Time), max(df$Browsing_Time), length.out = 100)
cats     <- levels(df$Category)

int_pred <- expand.grid(Browsing_Time = bt_range, Category = cats) %>%
  mutate(Predicted = predict(int_model, newdata = .))

ggplot(int_pred, aes(x = Browsing_Time, y = Predicted, color = Category)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("#E74C3C","#3498DB","#2ECC71","#F39C12","#9B59B6")) +
  labs(
    title    = "Interaction Plot: Browsing Time × Category on Purchase Amount",
    subtitle = "Non-parallel lines indicate interaction effects",
    x = "Browsing Time (min)", y = "Predicted Purchase Amount (USD)",
    color = "Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Insight:**  
> Non-parallel lines across categories indicate that the slope of Browsing_Time on
> Purchase_Amount differs by product category. For example, Electronics buyers may
> increase spending more steeply with browsing time than Books buyers, reflecting
> higher-consideration purchases. Parallel lines would indicate no interaction.

---

## 3.2 Coplots of Purchase_Amount vs. Browsing_Time by Category

```{r coplots}
ggplot(df, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.20, color = "#3498DB", size = 1) +
  geom_smooth(method = "lm", color = "#E74C3C", linewidth = 1,
              se = TRUE, fill = "#F1948A") +
  geom_smooth(method = "loess", color = "#27AE60", linewidth = 0.9,
              linetype = "dashed", se = FALSE) +
  facet_wrap(~ Category, ncol = 3, scales = "free_y") +
  labs(
    title    = "Coplot: Purchase Amount vs. Browsing Time by Category",
    subtitle = "Red = OLS | Green dashed = LOESS | Each panel = one Category",
    x = "Browsing Time (min)", y = "Purchase Amount (USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(face = "bold"),
    strip.text   = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "#EBF5FB", color = NA)
  )
```

> **Insight:**  
> Faceted coplots let us compare the local relationship within each category independently.
> Differences in slope, scatter spread, or LOESS curve shape across panels reveal
> category-specific purchasing patterns — e.g., whether high browsing always leads to
> higher spend in Electronics but not in Books.

---

## 3.3 Level Plot / Contour Plot: Browsing_Time, Number_of_Items, Purchase_Amount

```{r contour-plot}
# Create a 2D density grid using a binned average
contour_df <- df %>%
  mutate(
    BT_bin   = cut(Browsing_Time,    breaks = 20),
    Items_bin = cut(Number_of_Items, breaks = 10)
  ) %>%
  group_by(BT_bin, Items_bin) %>%
  summarise(
    Mean_Purchase  = mean(Purchase_Amount),
    BT_mid         = mean(Browsing_Time),
    Items_mid      = mean(Number_of_Items),
    .groups        = "drop"
  )

ggplot(contour_df, aes(x = BT_mid, y = Items_mid, fill = Mean_Purchase)) +
  geom_tile() +
  geom_contour(aes(z = Mean_Purchase), color = "white", alpha = 0.6, bins = 8) +
  scale_fill_gradient2(
    low      = "#2980B9",
    mid      = "#F39C12",
    high     = "#E74C3C",
    midpoint = median(contour_df$Mean_Purchase, na.rm = TRUE),
    labels   = dollar_format()
  ) +
  labs(
    title    = "Level Plot: Mean Purchase Amount by Browsing Time & Number of Items",
    subtitle = "Warmer colors = higher average purchase amount",
    x        = "Browsing Time (min)",
    y        = "Number of Items",
    fill     = "Mean Purchase\nAmount (USD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

```{r scatter-3d-proxy}
# 3-variable scatter: size encodes Purchase_Amount
ggplot(df %>% sample_n(500),
       aes(x = Browsing_Time, y = Number_of_Items,
           color = Purchase_Amount, size = Purchase_Amount)) +
  geom_point(alpha = 0.55) +
  scale_color_gradient(low = "#3498DB", high = "#E74C3C", labels = dollar_format()) +
  scale_size_continuous(range = c(0.5, 4), labels = dollar_format()) +
  labs(
    title    = "Trivariate Scatter: Browsing Time, Items, & Purchase Amount",
    subtitle = "Color and size both encode Purchase Amount (random sample of 500)",
    x = "Browsing Time (min)", y = "Number of Items",
    color = "Purchase\nAmount", size = "Purchase\nAmount"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Insight:**  
> The level/contour plot reveals joint regions of Browsing_Time and Number_of_Items
> that produce the highest average spending. High spending tends to emerge at specific
> combinations — for example, moderate browsing time with higher item counts, or
> long browsing with fewer, premium items (e.g., Electronics).

---

## 3.4 Multiple Regression & Model Selection

```{r multiple-regression}
# Full multiple regression model
mlr_full <- lm(
  Purchase_Amount ~ Browsing_Time + Number_of_Items + Satisfaction_Score,
  data = df
)

cat("=== Full Multiple Regression Summary ===\n")
summary(mlr_full)
```

```{r model-selection}
# Null model (intercept only)
mlr_null <- lm(Purchase_Amount ~ 1, data = df)

# Stepwise selection (both directions) using AIC
mlr_step <- step(mlr_null,
                 scope   = list(lower = mlr_null,
                                upper = mlr_full),
                 direction = "both",
                 trace   = 0)

cat("=== Stepwise Selected Model ===\n")
summary(mlr_step)

cat("\n=== Model Comparison (AIC) ===\n")
cat("Full model AIC:      ", round(AIC(mlr_full), 2), "\n")
cat("Stepwise model AIC:  ", round(AIC(mlr_step), 2), "\n")
cat("Null model AIC:      ", round(AIC(mlr_null), 2), "\n")
```

```{r variable-importance}
cat("=== Standardized Coefficients (Variable Importance) ===\n")
df_scaled <- df %>%
  mutate(across(c(Browsing_Time, Number_of_Items, Satisfaction_Score, Purchase_Amount),
                scale))

mlr_std <- lm(Purchase_Amount ~ Browsing_Time + Number_of_Items + Satisfaction_Score,
              data = df_scaled)

std_coefs <- tibble(
  Variable = names(coef(mlr_std))[-1],
  Std_Coef = round(coef(mlr_std)[-1], 4),
  Abs_Importance = abs(round(coef(mlr_std)[-1], 4))
) %>%
  arrange(desc(Abs_Importance))

print(std_coefs)
```

```{r importance-plot}
std_coefs %>%
  ggplot(aes(x = reorder(Variable, Abs_Importance), y = Std_Coef,
             fill = Std_Coef > 0)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = 0, linewidth = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2ECC71", "FALSE" = "#E74C3C")) +
  labs(
    title    = "Variable Importance: Standardized Regression Coefficients",
    subtitle = "Larger absolute value = greater importance to Purchase Amount",
    x = NULL, y = "Standardized Coefficient"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

```{r vif-check}
cat("=== Variance Inflation Factors (Multicollinearity Check) ===\n")
print(round(vif(mlr_full), 4))
```

```{r residual-diagnostics, fig.height=7}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(mlr_full, which = 1:4, col = "#3498DB", pch = 16, cex = 0.5)
par(mfrow = c(1, 1))
```

```{r final-comparison-table}
cat("=== Final Model Summary Table ===\n")
tibble(
  Model         = c("Null (intercept only)", "Simple Linear (Browsing_Time)",
                    "Multiple Regression (Full)", "Stepwise Selected"),
  R_Squared     = c(
    round(summary(mlr_null)$r.squared, 4),
    round(summary(lm_simple)$r.squared, 4),
    round(summary(mlr_full)$r.squared, 4),
    round(summary(mlr_step)$r.squared, 4)
  ),
  Adj_R_Squared = c(
    round(summary(mlr_null)$adj.r.squared, 4),
    round(summary(lm_simple)$adj.r.squared, 4),
    round(summary(mlr_full)$adj.r.squared, 4),
    round(summary(mlr_step)$adj.r.squared, 4)
  ),
  AIC           = c(
    round(AIC(mlr_null), 2),
    round(AIC(lm_simple), 2),
    round(AIC(mlr_full), 2),
    round(AIC(mlr_step), 2)
  )
)
```

> **Multiple Regression Interpretation:**  
>
> - **Browsing_Time:** A unit increase in browsing time corresponds to a β₁ USD change in
>   Purchase_Amount, holding other variables constant.  
> - **Number_of_Items:** Each additional item added to the cart is associated with a β₂ USD
>   change — likely positive (more items → higher total spend).  
> - **Satisfaction_Score:** Tests whether perceived satisfaction correlates with spending behavior.  
>
> **Model Selection:**  
> Stepwise AIC selection keeps only variables that improve model fit without unnecessary complexity.
> Variables removed by stepwise selection do not contribute significantly beyond what others already explain.
>
> **Variable Importance:**  
> Standardized coefficients put all predictors on the same scale — the predictor with the
> largest absolute standardized coefficient contributes most to predicting Purchase_Amount.
>
> **Multicollinearity (VIF):**  
> VIF values below 5 indicate acceptable multicollinearity. Values above 10 signal a
> problematic correlation between predictors that would inflate standard errors.

---

## Session Info

```{r session-info}
sessionInfo()
```
