---
title: "SA1 Data Mining"
author: "Awit, Julia Theressa and Deloyola, Jhen Reyniel"
date: "March 10, 2026`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
```

---

### Load Libraries

```{r load-libraries}
library(tidyverse)    
library(caret)       
library(glmnet)       
library(ggplot2)      
library(randomForest) 
library(rpart)        
library(rpart.plot)   
library(corrplot)     
library(scales)       
library(e1071)        
library(gridExtra)    

set.seed(42)
```

### Load Dataset

```{r load-data}
df <- read.csv("customer_churn.csv", stringsAsFactors = FALSE)

cat("=== Dataset Dimensions ===\n")
cat("Rows:", nrow(df), "| Columns:", ncol(df), "\n\n")

cat("=== Column Names ===\n")
print(names(df))

cat("\n=== Data Types ===\n")
str(df)
```

### Dataset Overview

```{r data-overview}
cat("=== Summary Statistics ===\n")
summary(df)

cat("\n=== Missing Values Per Column ===\n")
missing_vals <- colSums(is.na(df))
print(missing_vals[missing_vals > 0])
if (sum(missing_vals) == 0) cat("No missing values detected in initial scan.\n")

cat("\n=== Churn Distribution ===\n")
print(table(df$Churn))
print(prop.table(table(df$Churn)))
```

### Why Data Mining Is Important for This Dataset

> **Data Mining Importance:**  
> This dataset contains 10,000 customer records with demographics, service usage, and churn labels.
> The business problem — predicting which customers will churn — has direct financial impact.
> Retaining an existing customer is significantly cheaper than acquiring a new one.
>
> Data mining techniques allow us to:
> - **Discover patterns** hidden in high-dimensional customer data
> - **Build predictive models** to flag at-risk customers before they leave
> - **Identify key drivers** of churn (e.g., contract type, monthly charges)
> - **Enable proactive interventions** through customer segmentation and targeted offers
>
> Without data mining, the only strategy is reactive — waiting for a customer to leave before acting.

---

## Data Visualization

```{r viz-prep}
# Clean Churn and trim all character columns for safe plotting
df_plot <- df %>%
  mutate(across(where(is.character), trimws)) %>%
  mutate(Churn = ifelse(Churn %in% c("Yes", "1", 1), "Yes", "No"))
```

### Visualization 1

```{r viz-tenure}
df_plot %>%
  mutate(TenureGroup = cut(Tenure,
                           breaks = c(-Inf, 12, 24, 48, Inf),
                           labels = c("0-12 mo", "13-24 mo", "25-48 mo", "49+ mo"))) %>%
  group_by(TenureGroup) %>%
  summarise(
    Total     = n(),
    Churned   = sum(Churn == "Yes"),
    ChurnRate = Churned / Total,
    .groups   = "drop"
  ) %>%
  ggplot(aes(x = TenureGroup, y = ChurnRate, fill = TenureGroup)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = percent(ChurnRate, accuracy = 0.1)),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.65)) +
  scale_fill_manual(values = c("#E74C3C", "#E67E22", "#3498DB", "#2ECC71")) +
  labs(
    title    = "Churn Rate by Customer Tenure Group",
    subtitle = "Newer customers churn at a significantly higher rate",
    x = "Tenure Group", y = "Churn Rate (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

**Insight:** Customers with less than 12 months of tenure have the highest churn rate. As tenure increases, churn rate drops substantially — indicating that long-term customers are more loyal and less likely to leave.

---

### Visualization 2

```{r viz-charges}
ggplot(df_plot, aes(x = MonthlyCharges, fill = Churn)) +
  geom_density(alpha = 0.55, linewidth = 0.8) +
  scale_fill_manual(values = c("No" = "#2ECC71", "Yes" = "#E74C3C")) +
  labs(
    title    = "Monthly Charges Distribution by Churn Status",
    subtitle = "Churned customers tend to pay higher monthly charges",
    x = "Monthly Charges ($)", y = "Density", fill = "Churned"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

**Insight:** Churned customers cluster at higher monthly charges (~$70–$100+), while retained customers spread toward lower charges. High billing may indicate price sensitivity as a churn trigger.

---

### Visualization 3

```{r viz-contract}
df_plot %>%
  filter(Contract != "") %>%
  group_by(Contract, Churn) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Contract) %>%
  mutate(Pct = Count / sum(Count)) %>%
  ggplot(aes(x = Contract, y = Pct, fill = Churn)) +
  geom_col(position = "fill", width = 0.55) +
  geom_text(aes(label = percent(Pct, accuracy = 1)),
            position = position_fill(vjust = 0.5),
            size = 4.2, color = "white", fontface = "bold") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("No" = "#2980B9", "Yes" = "#E74C3C")) +
  labs(
    title    = "Churn Rate by Contract Type",
    subtitle = "Month-to-month contracts have the highest churn risk",
    x = "Contract Type", y = "Proportion", fill = "Churned"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

**Insight:** Month-to-month contract holders show dramatically higher churn rates. Long-term contract customers are effectively "locked in" — contract type is one of the strongest predictors of churn.

---

### Visualization 4

```{r viz-corr}
num_vars <- df_plot %>%
  mutate(
    ChurnBinary   = ifelse(Churn == "Yes", 1, 0),
    SeniorCitizen = as.numeric(SeniorCitizen)
  ) %>%
  select(Tenure, MonthlyCharges, TotalCharges, SeniorCitizen, ChurnBinary) %>%
  drop_na()

corrplot(cor(num_vars, use = "complete.obs"),
         method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         col = colorRampPalette(c("#2980B9", "white", "#E74C3C"))(200),
         title = "Correlation Matrix of Numerical Features",
         mar   = c(0, 0, 2, 0))
```

**Insight:** `TotalCharges` and `Tenure` are highly correlated (total charges accumulate over time). `MonthlyCharges` shows a moderate positive correlation with churn, confirming the density plot finding.

---

## Data Transformation

```{r data-transform}
df_clean <- df %>%
  # Trim all whitespace from character columns
  mutate(across(where(is.character), trimws)) %>%

  # Replace empty strings with NA
  mutate(across(where(is.character), ~na_if(., ""))) %>%

  # Standardize Churn to Yes/No factor
  mutate(Churn = ifelse(Churn %in% c("Yes", "1", 1), "Yes", "No"),
         Churn = factor(Churn, levels = c("No", "Yes"))) %>%

  # Coerce numeric columns (TotalCharges sometimes read as character)
  mutate(
    Tenure         = as.numeric(Tenure),
    MonthlyCharges = as.numeric(MonthlyCharges),
    TotalCharges   = as.numeric(TotalCharges)
  ) %>%

  # Impute numeric NAs with median
  mutate(
    Tenure         = ifelse(is.na(Tenure),         median(Tenure,         na.rm = TRUE), Tenure),
    MonthlyCharges = ifelse(is.na(MonthlyCharges), median(MonthlyCharges, na.rm = TRUE), MonthlyCharges),
    TotalCharges   = ifelse(is.na(TotalCharges),   median(TotalCharges,   na.rm = TRUE), TotalCharges)
  ) %>%

  # Convert categoricals to factors
  mutate(
    Gender          = factor(Gender),
    SeniorCitizen   = factor(as.character(SeniorCitizen),
                             levels = c("0","1"), labels = c("No","Yes")),
    Partner         = factor(Partner),
    Dependents      = factor(Dependents),
    PhoneService    = factor(PhoneService),
    InternetService = factor(InternetService),
    Contract        = factor(Contract)
  ) %>%

  # Drop rows that still have NA (e.g., missing categoricals)
  drop_na() %>%

  # Drop unused factor levels
  droplevels()

cat("=== Post-Transformation Structure ===\n")
str(df_clean)

cat("\n=== Remaining Missing Values ===\n")
print(colSums(is.na(df_clean)))

cat("\nRows remaining:", nrow(df_clean), "\n")
```

### Normalize Numerical Features

```{r normalize}
df_clean <- df_clean %>%
  mutate(
    Tenure_scaled         = as.numeric(scale(Tenure)),
    MonthlyCharges_scaled = as.numeric(scale(MonthlyCharges)),
    TotalCharges_scaled   = as.numeric(scale(TotalCharges))
  )

cat("=== Scaled Feature Summary ===\n")
summary(df_clean %>% select(Tenure_scaled, MonthlyCharges_scaled, TotalCharges_scaled))
```

---

### Remove Outliers (IQR Method)

```{r outliers}
remove_outliers_iqr <- function(data, col) {
  q1    <- quantile(data[[col]], 0.25, na.rm = TRUE)
  q3    <- quantile(data[[col]], 0.75, na.rm = TRUE)
  iqr   <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  data %>% filter(.data[[col]] >= lower & .data[[col]] <= upper)
}

nrow_before <- nrow(df_clean)
df_clean    <- df_clean %>%
  remove_outliers_iqr("MonthlyCharges") %>%
  remove_outliers_iqr("TotalCharges")
nrow_after  <- nrow(df_clean)

cat("Rows before outlier removal:", nrow_before, "\n")
cat("Rows after  outlier removal:", nrow_after,  "\n")
cat("Outliers removed:           ", nrow_before - nrow_after, "\n")
```

### Create Derived Variables

```{r derived-vars}
df_clean <- df_clean %>%
  mutate(
    AvgMonthlySpend = ifelse(Tenure > 0, TotalCharges / Tenure, MonthlyCharges),
    TenureGroup     = cut(Tenure,
                          breaks = c(-Inf, 12, 24, 48, Inf),
                          labels = c("New", "Early", "Mid", "Long-term")),
    HighValue       = factor(ifelse(MonthlyCharges > median(MonthlyCharges), "Yes", "No"))
  )

cat("=== Derived Variables Preview ===\n")
df_clean %>%
  select(Tenure, TotalCharges, AvgMonthlySpend, TenureGroup, HighValue) %>%
  head(8)
```

### Aggregate Summary by Contract and Churn

```{r aggregation}
df_clean %>%
  group_by(Contract, Churn) %>%
  summarise(
    Count       = n(),
    Avg_Tenure  = round(mean(Tenure), 1),
    Avg_Monthly = round(mean(MonthlyCharges), 2),
    Avg_Total   = round(mean(TotalCharges), 2),
    .groups     = "drop"
  ) %>%
  arrange(Contract, Churn)
```

---

## EDA Summary & Key Takeaways

> **Key Takeaways:**
>
> 1. **Tenure is inversely related to churn** — the first 12 months are the highest-risk window.
> 2. **Higher monthly charges correlate with churn** — price sensitivity is evident.
> 3. **Contract type is a strong predictor** — month-to-month customers churn far more.
> 4. **TotalCharges and Tenure are multicollinear** — regularization (Ridge/Lasso) or PCA recommended.
> 5. **Senior citizens (~15%)** may have distinct churn behavior warranting separate analysis.

---

# Tuning Predictive Models

## Prepare Modeling Dataset

```{r model-prep}
# ── Build clean modeling dataframe ─────────────────────────────────────────
model_df <- df_clean %>%
  select(Churn, Tenure, MonthlyCharges, TotalCharges,
         SeniorCitizen, Contract, InternetService,
         Gender, Partner, Dependents, PhoneService) %>%

  # Trim + remove empty strings one more time to be safe
  mutate(across(where(is.character), trimws)) %>%
  mutate(across(where(is.character), ~na_if(., ""))) %>%

  # Ensure correct types
  mutate(
    Tenure         = as.numeric(Tenure),
    MonthlyCharges = as.numeric(MonthlyCharges),
    TotalCharges   = as.numeric(TotalCharges),
    Churn           = factor(Churn, levels = c("No", "Yes")),
    Contract        = factor(Contract),
    InternetService = factor(InternetService),
    SeniorCitizen   = factor(as.character(SeniorCitizen)),
    Gender          = factor(Gender),
    Partner         = factor(Partner),
    Dependents      = factor(Dependents),
    PhoneService    = factor(PhoneService)
  ) %>%
  drop_na() %>%
  droplevels()

# Verify — must show Factor/num, NO chr
cat("=== Model DataFrame Structure ===\n")
str(model_df)

cat("\n=== Churn Distribution ===\n")
print(prop.table(table(model_df$Churn)))

# ── Train / Test split (70 / 30) ───────────────────────────────────────────
set.seed(42)
train_idx  <- createDataPartition(model_df$Churn, p = 0.70, list = FALSE)
train_data <- model_df[train_idx, ]
test_data  <- model_df[-train_idx, ]

cat("\nTrain size:", nrow(train_data), "| Test size:", nrow(test_data), "\n")
```

---

## Decision Tree vs Logistic Regression

### Decision Tree

```{r decision-tree}
dt_model <- rpart(
  Churn ~ Tenure + MonthlyCharges + TotalCharges +
          Contract + InternetService + SeniorCitizen,
  data   = train_data,
  method = "class",
  control = rpart.control(
    cp        = 0.001,
    maxdepth  = 5,
    minsplit  = 20,
    minbucket = 7
  )
)

cat("=== Decision Tree CP Table ===\n")
printcp(dt_model)

# Prune to best cp
best_cp  <- dt_model$cptable[which.min(dt_model$cptable[, "xerror"]), "CP"]
dt_model <- prune(dt_model, cp = best_cp)

rpart.plot(dt_model,
           type        = 4,
           extra       = 106,
           box.palette = c("#E74C3C", "#2ECC71"),
           fallen.leaves = TRUE,
           main        = "Decision Tree: Customer Churn Prediction")

# Performance
dt_pred <- predict(dt_model, test_data, type = "class")
dt_cm   <- confusionMatrix(dt_pred, test_data$Churn, positive = "Yes")
cat("\n=== Decision Tree Test Performance ===\n")
print(dt_cm$overall[c("Accuracy", "Kappa")])
print(dt_cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])
```

### Logistic Regression

```{r logistic-regression-complexity}
lr_model <- glm(
  Churn ~ Tenure + MonthlyCharges + TotalCharges +
          Contract + InternetService + SeniorCitizen,
  data   = train_data,
  family = binomial(link = "logit")
)

lr_prob <- predict(lr_model, test_data, type = "response")
lr_pred <- factor(ifelse(lr_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
lr_cm   <- confusionMatrix(lr_pred, test_data$Churn, positive = "Yes")

cat("=== Logistic Regression Test Performance ===\n")
print(lr_cm$overall[c("Accuracy", "Kappa")])
print(lr_cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])
```

### Complexity Comparison Table

```{r complexity-compare}
tibble(
  Aspect               = c("Model Type", "Interpretability",
                            "Handles Non-linearity", "Overfitting Risk",
                            "Test Accuracy"),
  Decision_Tree        = c("Non-parametric", "High (visual splits)",
                            "Yes", "High without pruning",
                            round(dt_cm$overall["Accuracy"], 4)),
  Logistic_Regression  = c("Parametric", "High (coefficients)",
                            "No (linear boundary)", "Low",
                            round(lr_cm$overall["Accuracy"], 4))
)
```

> **Trade-off Discussion:**  
> Decision trees capture non-linear interactions and are easy to visualize, but overfit without pruning.
> Logistic Regression is stable and interpretable via odds ratios, but assumes linearity.
> For churn prediction with mixed variable types, ensemble methods (Random Forest) generally outperform both.

---

## 2.2 Bias-Variance Trade-Off

```{r bias-variance, fig.height=4.5}
depths    <- 1:12
acc_train <- numeric(length(depths))
acc_test  <- numeric(length(depths))

for (i in seq_along(depths)) {
  m <- rpart(
    Churn ~ Tenure + MonthlyCharges + TotalCharges + Contract + InternetService,
    data    = train_data,
    method  = "class",
    control = rpart.control(maxdepth = depths[i], cp = 0)
  )
  acc_train[i] <- mean(predict(m, train_data, type = "class") == train_data$Churn)
  acc_test[i]  <- mean(predict(m, test_data,  type = "class") == test_data$Churn)
}

tibble(Depth = depths, Train = acc_train, Test = acc_test) %>%
  pivot_longer(c(Train, Test), names_to = "Set", values_to = "Accuracy") %>%
  ggplot(aes(x = Depth, y = Accuracy, color = Set, group = Set)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Train" = "#3498DB", "Test" = "#E74C3C")) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = depths) +
  labs(
    title    = "Bias-Variance Trade-off: Tree Depth vs. Accuracy",
    subtitle = "Train accuracy rises with depth; test accuracy plateaus or drops",
    x = "Tree Max Depth", y = "Accuracy", color = "Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

> **Explanation:**  
> - **High Bias (depth 1–2):** Model is too simple; underfits both train and test.  
> - **High Variance (depth 8+):** Model memorizes training noise; test accuracy degrades.  
> - **Optimal zone (~depth 4–6):** Best generalization — balanced bias and variance.  
> - Logistic Regression has fixed high bias (linear boundary) but consistently low variance.

---

## Cross-Validation (k = 10)

```{r cross-validation}
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# Decision Tree (CV)
dt_cv <- train(
  Churn ~ Tenure + MonthlyCharges + TotalCharges + Contract + InternetService + SeniorCitizen,
  data       = train_data,
  method     = "rpart",
  metric     = "ROC",
  trControl  = cv_ctrl,
  tuneLength = 10
)

# Logistic Regression (CV)
lr_cv <- train(
  Churn ~ Tenure + MonthlyCharges + TotalCharges + Contract + InternetService + SeniorCitizen,
  data      = train_data,
  method    = "glm",
  family    = "binomial",
  metric    = "ROC",
  trControl = cv_ctrl
)

cat("=== 10-Fold CV ROC Scores ===\n")
cat("Decision Tree  (best ROC):", round(max(dt_cv$results$ROC), 4), "\n")
cat("Logistic Regression (ROC):", round(lr_cv$results$ROC, 4), "\n")
```

```{r cv-metrics}
# Helper to extract metrics from CV predictions
get_cv_metrics <- function(preds_df, model_name) {
  p <- preds_df %>%
    mutate(pred = factor(pred, levels = c("No","Yes")),
           obs  = factor(obs,  levels = c("No","Yes")))
  cm <- confusionMatrix(p$pred, p$obs, positive = "Yes")
  tibble(
    Model     = model_name,
    Accuracy  = round(cm$overall["Accuracy"],    4),
    Precision = round(cm$byClass["Precision"],   4),
    Recall    = round(cm$byClass["Sensitivity"], 4),
    F1        = round(cm$byClass["F1"],          4)
  )
}

dt_preds_cv <- dt_cv$pred %>%
  filter(cp == dt_cv$bestTune$cp)

lr_preds_cv <- lr_cv$pred

cv_results <- bind_rows(
  get_cv_metrics(dt_preds_cv, "Decision Tree"),
  get_cv_metrics(lr_preds_cv, "Logistic Regression")
)

cat("=== 10-Fold CV Classification Metrics ===\n")
print(cv_results)
```

> **Interpretation:**  
> 10-fold CV gives a more reliable estimate than a single split. F1-score is most informative here due to class imbalance — a model that predicts "No" for everyone achieves ~73% accuracy but zero recall for churners.

---

## Random Forest with Hyperparameter Tuning

```{r random-forest}
rf_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary
)

rf_grid <- expand.grid(mtry = c(2, 3, 4, 5, 6))

rf_model <- train(
  Churn ~ Tenure + MonthlyCharges + TotalCharges + Contract +
          InternetService + SeniorCitizen + Gender + Partner + Dependents,
  data      = train_data,
  method    = "rf",
  metric    = "ROC",
  trControl = rf_ctrl,
  tuneGrid  = rf_grid,
  ntree     = 200
)

cat("=== Random Forest Best Tune ===\n")
print(rf_model$bestTune)

cat("\n=== CV Results by mtry ===\n")
print(rf_model$results[, c("mtry", "ROC", "Sens", "Spec")])
```

```{r rf-performance}
rf_pred <- predict(rf_model, test_data)
rf_cm   <- confusionMatrix(rf_pred, test_data$Churn, positive = "Yes")

cat("=== Random Forest Final Performance ===\n")
print(rf_cm$overall[c("Accuracy", "Kappa")])
print(rf_cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])

# Variable importance
varImpPlot(rf_model$finalModel,
           n.var = 10,
           main  = "Random Forest: Top 10 Variable Importance",
           col   = "#2C3E50", pch = 16)
```

> **Summary:** Random Forest ensembles many decorrelated trees and significantly outperforms a single tree. Grid search over `mtry` selects the optimal feature subset per split. `Contract`, `Tenure`, and `MonthlyCharges` typically emerge as the most important predictors.

---

# Regression-Based Methods

## 3.1 Logistic Regression (Interpretation)

```{r logistic-regression-unit3}
lr_unit3 <- glm(
  Churn ~ Tenure + MonthlyCharges + TotalCharges,
  data   = train_data,
  family = binomial(link = "logit")
)

cat("=== Logistic Regression Summary ===\n")
summary(lr_unit3)
```

```{r lr-odds-ratios}
cat("=== Odds Ratios with 95% Confidence Intervals ===\n")
or_table <- exp(cbind(OddsRatio = coef(lr_unit3), confint(lr_unit3)))
print(round(or_table, 4))

lr3_prob <- predict(lr_unit3, test_data, type = "response")
lr3_pred <- factor(ifelse(lr3_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
lr3_cm   <- confusionMatrix(lr3_pred, test_data$Churn, positive = "Yes")

cat("\nTest Accuracy:", round(lr3_cm$overall["Accuracy"], 4), "\n")
cat("AIC:          ", round(AIC(lr_unit3), 2), "\n")
```

> **Coefficient Interpretation:**  
> - **Tenure (negative):** Longer-tenured customers have lower log-odds of churning — loyalty grows over time.  
> - **MonthlyCharges (positive):** Higher bills increase the probability of churn — price sensitivity is significant.  
> - **TotalCharges:** Multicollinear with Tenure; interpret with caution alongside the other two.  
> - **P-values < 0.05** confirm statistical significance of predictors.

---

## High-Dimensional Regression & PCA

```{r pca}
pca_data <- df_clean %>%
  select(Tenure, MonthlyCharges, TotalCharges) %>%
  drop_na() %>%
  scale()

pca_result <- prcomp(pca_data, center = FALSE, scale. = FALSE)

cat("=== PCA Summary ===\n")
summary(pca_result)

# Scree plot
var_explained <- (pca_result$sdev^2) / sum(pca_result$sdev^2)
tibble(
  PC    = paste0("PC", seq_along(var_explained)),
  Var   = var_explained,
  Cumul = cumsum(var_explained)
) %>%
  ggplot(aes(x = PC)) +
  geom_col(aes(y = Var), fill = "#3498DB", width = 0.5) +
  geom_line(aes(y = Cumul, group = 1), color = "#E74C3C", linewidth = 1.2) +
  geom_point(aes(y = Cumul), color = "#E74C3C", size = 3) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray50") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "PCA Scree Plot: Variance Explained",
    subtitle = "Red = cumulative variance; dashed = 95% threshold",
    x = "Principal Component", y = "Variance Explained"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

cat("\n=== PCA Loadings ===\n")
print(round(pca_result$rotation, 4))
```

> **Discussion:**  
> High-dimensional regression challenges include multicollinearity (inflated SE), overfitting, and unstable coefficient estimates. Solutions: **PCA** produces orthogonal components that eliminate multicollinearity; **Ridge** shrinks all coefficients; **Lasso** zeros out irrelevant ones. PC1 here primarily captures the "customer lifetime value" axis (Tenure + TotalCharges), while PC2 isolates price level (MonthlyCharges).

---

## Ridge Regression

```{r ridge-prep}
# Build model matrices — factors must have clean levels (ensured by droplevels() above)
X_train <- model.matrix(
  ~ Tenure + MonthlyCharges + TotalCharges +
    SeniorCitizen + Gender + Partner + Dependents +
    Contract + InternetService,
  data = train_data
)[, -1]   # drop intercept column

X_test <- model.matrix(
  ~ Tenure + MonthlyCharges + TotalCharges +
    SeniorCitizen + Gender + Partner + Dependents +
    Contract + InternetService,
  data = test_data
)[, -1]

y_train <- ifelse(train_data$Churn == "Yes", 1, 0)
y_test  <- ifelse(test_data$Churn  == "Yes", 1, 0)

cat("X_train dimensions:", dim(X_train), "\n")
cat("Column names:\n")
print(colnames(X_train))
```

```{r ridge-regression}
# Cross-validated Ridge (alpha = 0)
ridge_cv <- cv.glmnet(X_train, y_train,
                      family       = "binomial",
                      alpha        = 0,
                      nfolds       = 10,
                      type.measure = "class")

cat("=== Ridge: Optimal Lambda ===\n")
cat("Lambda min:", round(ridge_cv$lambda.min, 6), "\n")
cat("Lambda 1se:", round(ridge_cv$lambda.1se, 6), "\n")

plot(ridge_cv, main = "Ridge: CV Error vs. log(Lambda)")
```

```{r ridge-fit}
ridge_final <- glmnet(X_train, y_train,
                      family = "binomial",
                      alpha  = 0,
                      lambda = ridge_cv$lambda.min)

cat("=== Ridge Coefficients (lambda.min) ===\n")
print(round(as.matrix(coef(ridge_final)), 4))

ridge_prob <- predict(ridge_final, newx = X_test, type = "response")
ridge_pred <- factor(ifelse(ridge_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
ridge_cm   <- confusionMatrix(ridge_pred, test_data$Churn, positive = "Yes")

cat("\n=== Ridge Test Performance ===\n")
print(ridge_cm$overall[c("Accuracy", "Kappa")])
print(ridge_cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])
```

> **Ridge Summary:** The L2 penalty (λΣβ²) shrinks all coefficients toward zero but retains every predictor. Optimal λ is chosen by 10-fold CV. Ridge is ideal here because of the multicollinearity between Tenure and TotalCharges — it stabilizes estimates without discarding variables.

---

## Lasso Regression

```{r lasso-regression}
# Cross-validated Lasso (alpha = 1)
lasso_cv <- cv.glmnet(X_train, y_train,
                      family       = "binomial",
                      alpha        = 1,
                      nfolds       = 10,
                      type.measure = "class")

cat("=== Lasso: Optimal Lambda ===\n")
cat("Lambda min:", round(lasso_cv$lambda.min, 6), "\n")
cat("Lambda 1se:", round(lasso_cv$lambda.1se, 6), "\n")

plot(lasso_cv, main = "Lasso: CV Error vs. log(Lambda)")
```

```{r lasso-fit}
lasso_final <- glmnet(X_train, y_train,
                      family = "binomial",
                      alpha  = 1,
                      lambda = lasso_cv$lambda.min)

lasso_coefs <- as.matrix(coef(lasso_final))

cat("=== Lasso Coefficients (lambda.min) ===\n")
print(round(lasso_coefs, 4))

cat("\n=== Features Zeroed Out by Lasso ===\n")
zeroed <- rownames(lasso_coefs)[lasso_coefs[, 1] == 0]
if (length(zeroed) == 0) cat("None — all features retained at lambda.min\n") else cat(paste(zeroed, collapse = ", "), "\n")

cat("\n=== Active Features Retained ===\n")
active <- rownames(lasso_coefs)[lasso_coefs[, 1] != 0]
cat(paste(active, collapse = ", "), "\n")

lasso_prob <- predict(lasso_final, newx = X_test, type = "response")
lasso_pred <- factor(ifelse(lasso_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
lasso_cm   <- confusionMatrix(lasso_pred, test_data$Churn, positive = "Yes")

cat("\n=== Lasso Test Performance ===\n")
print(lasso_cm$overall[c("Accuracy", "Kappa")])
print(lasso_cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])
```

> **Lasso Benefits:** The L1 penalty (λΣ|β|) forces some coefficients exactly to zero, performing automatic feature selection. This produces a sparse, interpretable model. Lasso likely retains Contract, Tenure, and MonthlyCharges while zeroing out less informative demographics — confirming their importance as churn drivers.

---

## Final Model Comparison

```{r final-comparison}
results <- tibble(
  Model     = c("Decision Tree", "Logistic Regression",
                "Random Forest", "Ridge Regression", "Lasso Regression"),
  Accuracy  = c(round(dt_cm$overall["Accuracy"],    4),
                round(lr_cm$overall["Accuracy"],    4),
                round(rf_cm$overall["Accuracy"],    4),
                round(ridge_cm$overall["Accuracy"], 4),
                round(lasso_cm$overall["Accuracy"], 4)),
  F1_Score  = c(round(dt_cm$byClass["F1"],    4),
                round(lr_cm$byClass["F1"],    4),
                round(rf_cm$byClass["F1"],    4),
                round(ridge_cm$byClass["F1"], 4),
                round(lasso_cm$byClass["F1"], 4)),
  Recall    = c(round(dt_cm$byClass["Sensitivity"],    4),
                round(lr_cm$byClass["Sensitivity"],    4),
                round(rf_cm$byClass["Sensitivity"],    4),
                round(ridge_cm$byClass["Sensitivity"], 4),
                round(lasso_cm$byClass["Sensitivity"], 4))
)

cat("=== Final Model Comparison Table ===\n")
print(results)

# Bar chart comparison
results %>%
  pivot_longer(c(Accuracy, F1_Score, Recall),
               names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = reorder(Model, Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge", width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("#3498DB", "#E74C3C", "#2ECC71")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.05)) +
  labs(
    title    = "Final Model Performance Comparison",
    subtitle = "Accuracy, F1-Score, and Recall across all models",
    x = NULL, y = "Score", fill = "Metric"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Session Info

```{r session-info}
sessionInfo()
```
